Alien species

### 2023-03-14 DRAFT_alien_species.Rmd
write intro

# (PART\*) INDICATORS{-}

# Slitasje og kjørespor

@@

## Introduction
This indicator represent human caused damage or **disturbance to soils and vegetation**, typically from things like recreational activities and off-road vehicle traffic. Effects from domestic animal is not included here, and the upper reference value is 0% disturbance. The data come from a combination non-random field surveys and area-representative surveys.

>>>>

---
title: "Alien species"
output:
  html_document
---
# Alien species {#alien-species}

@@

## Introduction
This indicator represent the the relative area with an absence of alien species. It is in a way the flip-side of _alien species abundance_. The later can perhaps be classed as a pressure variable, whilst the former reacts as direct and linear consequence of changes to this variable.

A related variable has been used previously for [alpine](https://ninanor.github.io/IBECA/faktaark.html#Frav%C3%A6r_av_fremmede_arter) and forest ecosystems in Norway. These indicators were based on ANO, a national monitoring program, where alien species cover was estimated to the closest whole %. The upper and lower reference levels that was used back then was 0 or 100% of the species composition made up of alien species, respectively. The threshold value between good and reduced condition was defined as 5%.

This time we are using the [nature type mapping dataset](#naturtype) which records the _effect_ of alien species,, rather than the % cover. The recording scale is called R7 (Fig. \@ref(fig:R7)) and is an ordinal scale which has a non-linear relationship with %cover recorded in ANO (Fig. \@ref(fig:fremmedart-prosentVSinnslag)). In order to combine these two data sets, we must convert one of these scales to match the other.

Here I base the indicator on the R7 scale, since this is the bigger data set. I therfore convert %cover into R7 classes based on what I can interpret from Fig. \@ref(fig:fremmedart-prosentVSinnslag).

```{r R7, fig.cap="The R7 scale used when recording the effect of alien species in the nature type monitoring program"}
knitr::include_graphics("images/R7 scale.PNG")
```

```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring progra with %cover (y-axis), which is how alien species is recorded in ANO."}
knitr::include_graphics("images/fremmedart_prosentVSinnslag.PNG")
```
